---
title: "Practical Machine Learning Course Project"
author: "Brian Altman"
date: "January 28, 2016"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

###Background
With devices such as Jawbone Up, Nike FuelBand, and Fitbit it is possible
to collect a large amount of data about personal activity relatively inexpensively. 
These type of devices are part of the quantified self movement - a group of 
enthusiasts who take measurements about themselves regularly to improve their health, 
to find patterns in their behavior, or because they are tech geeks. One thing that
people regularly do is quantify how much of a particular activity they do, but 
they rarely quantify how well they do it. 

In this project, your goal will be to use data from accelerometers on the 
belt, forearm, arm, and dumbell of 6 participants. They were asked to perform 
barbell lifts correctly and incorrectly in 5 different ways:


* A: according to the specification 
* B: elbows to the front
* C: lift dumbbell only halfway 
* D: lower dumbbell only halfway
* E: hips to the fron

Data
The training data for this project are available here:
      https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
      https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

More information is available from the #website here: http://groupware.les.inf.puc-rio.br/har 
(see the section on the Weight Lifting Exercise Dataset).


####Libraries and packages

```{r libraries}
library(data.table)
library(caret)
#install.packages("randomForest")
#install.packages("e1071")
knitr::opts_chunk$set(cache=TRUE)
```

####Load data sets
```{r load}
pml_training <- fread("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
pml_testing  <- fread("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
pml_training <- as.data.frame(pml_training)
```

####Data Preparation
Remove any columns that have missing values
```{r DataPrep1}
MissingDataColumns <- sapply(pml_training, function (x) any(is.na(x) | x == " " | x == ""))
pml_training_without_missing <- pml_training[,!MissingDataColumns]
```

Remove columns unrelated to sensors of interest
```{r DataPrep2}
TrainingSensorColumns <- grep(pattern = "_belt|_arm|_dumbbell|_forearm", names(pml_training_without_missing))
pml_training_cleaned <- pml_training_without_missing[,c(TrainingSensorColumns,60)]
pml_training_cleaned$classe <- as.factor(pml_training_cleaned$classe)
str(pml_training_cleaned)
```

Remove values close to zero
```{r DataPrep3}
pml_training_cleaned_pp <-preProcess(pml_training_cleaned,method=c('center', 'scale'))
pml_training_cleaned_pp_pr <- predict(pml_training_cleaned_pp, pml_training_cleaned)
pml_training_cleaned_pp_pr$classe <- pml_training_cleaned$classe
NearZeroValues <- nearZeroVar(pml_training_cleaned_pp_pr,saveMetrics=TRUE)
pml_training_cleaned_pp_pr_z <- pml_training_cleaned_pp_pr[,NearZeroValues$nzv==FALSE]
```

####Partition the training cleaned data set into training and validation
```{r Partition}
pml_training_partition <- createDataPartition(pml_training_cleaned_pp_pr_z$classe, p=0.75)
pml_training_cleaned_Train      <- pml_training_cleaned[ pml_training_partition  [[1]],]
pml_training_cleaned_Validaton  <- pml_training_cleaned[-pml_training_partition  [[1]],]
dim(pml_training_cleaned_Train)
dim(pml_training_cleaned_Validaton)
```


####Using Random forest, train the model using the training partition
```{r ModelTrain}
#Sys.time()  #used to monitor progress
#TC1 <- trainControl(method='cv', allowParallel=TRUE, number=5,verboseIter = TRUE)  #used to monitor progress
TC1 <- trainControl(method='cv', allowParallel=TRUE, number=5)
TrainingModel <- train(classe ~., method="rf", data=pml_training_cleaned_Train,trControl=TC1)
TrainingModel
#Sys.time() #used to monitor progress
```

####Determine accuracy of model
```{r Accuracy}
TrainingPrediction <- predict(TrainingModel, pml_training_cleaned_Train)
confusionMatrix(TrainingPrediction, pml_training_cleaned_Train$classe)

trainingPred <- predict(TrainingModel, pml_training_cleaned_Validaton)
confusionMatrix(trainingPred, pml_training_cleaned_Validaton$classe)
```

####Based on pml_testing data, create predictions for class answers
```{r Predictions}
pml_testing_prediction <- predict(TrainingModel, pml_testing)
pml_testing_prediction
```
